\chapter{Conclusion}

\par In conclusion, the performance, data analysis, and scalability of \textbf{SQLite} databases are poor, but they are extremely convenient and require little setup. They may be used as mirror databases to analyze data copied from another DBMS without impacting the performance of the primary database instance. \textbf{PostgreSQL} and other advanced SQL implementations are efficient at storing and analyzing data. Still, they do not have any ability to alert the user about values outside allowed ranges, requiring external services like Grafana. \textbf{Apache Cassandra} is a flexible DBMS that allows for easy vertical scalability, making it useful for large energy clusters or those that may grow in size in the future. MongoDB is also scalable and has a flexible data schema, but its document-oriented schema does not reflect well the nature of data gathered by energy clusters.
\par \textbf{Prometheus} seems to be the ideal candidate for use in energy clusters. Its inflexible schema accurately reflects the data gathered by energy meters, and its pull-based model allows for monitoring the health of the connection between the database and every meter. Its query language allows for monitoring data falling outside the norm and does not require any external service to analyze and visualize data. However, Prometheus' capabilities can be extended with a service like Grafana. When high-end equipment is unavailable, data may be sent in bigger batches and at higher intervals to reduce the lag caused by network latency.
\par The data performance analysis was performed on a single local machine with single 1-node instances of each DBMS. A more extensive study with multiple instances/nodes and multiple data sources may be conducted to assess how various DBMSs react in such an environment. \textbf{Graph} databases, such as neo4j, were not considered in the comparison but may also be considered in future research. The \textbf{NewSQL} paradigm, which combines the SQL model with the scalability of NoSQL approaches, is also worth consideration.